On the properties of discs around accreting brown dwarfs

### 2010-06-11 BD_disc/paper.tex
Final version to be submitted
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2297 25464f21-6099-402c-b30a-640f1d40d38b

In surveys, particularly those attempting to discern the PMS disc
fraction this disentangling usually takes place using broad-band
photometric measures and cuts in colour-colour or colour-magnitude
space \citep{luhman_2005b,luhman_2008,gutermuth_2008}.  The important
quantity here is the separation in wavelength between the emission
peaks for the stellar and thermal disc components, as it is these
components which must be isolated. For BDD systems it is likely that
difficulties detecting the disc will be exacerbated by the lower
temperatures of the photosphere and therefore the smaller separation
in wavelength from the thermal disc emission component, compared to
CTTS systems. Therefore, to derive disc fractions for BD populations
we require detailed comparison models with which to guide disc
candidate selection.

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be indepedent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}: At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photopsheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may
well be a population of low-accretion-rate CTTS which are current
missing from the surveys.

@@

In summary the key required input variables to setup the model grid
are as follows: Stellar age and mass (which are used to derive the
stellar flux) and accretion rate, areal coverage and rotation period
(which are used to determine the the accretion flux and temperature).
The disc parameters are set by a disc mass (expressed as a fraction of
the BD mass), and a power-law distribution for the disc surface
density.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a
swift convergence of the sublimated rim. The dust sublimation proceeds
as follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations.

@@

The disc surface density is conserved in the initial density
structure, $\Sigma ^{\beta - \alpha}$. The initial scaleheight ($h$)
and density ($\rho$) follow $h\propto r^{\beta}$
($h=h_0(r/R_*)^{\beta}$) and $\rho \propto r^{-\alpha}$ ($\rho
=\rho_0\frac{R_*}{r}exp(-\frac{1}{2}[z/h(r)]^2)$) respectively., where
$r$ and $z$ are the radial and vertical coordinates. The initial
values of $\alpha$ and $\beta$ were 2.1 and 1.1 respectively. The
values for $\alpha$ and $\beta$ were chosen to optimise resolution of
the vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed at the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{walker_2004}. The initial disc scaleheight at 100\,AU, $h(100)$,
was set to 25\,AU. As the simulation used vertical hydrostatic
equilibrium and dust sublimation, both the disc scaleheight and inner
edge location then evolved in the systems dependent on the input
parameters. In this section we discuss the structure of the discs in
terms of these two generated characteristics, i.e the disc scaleheight
and inner edge location.

@@

The resulting converged disc structures, as discussed in Section
\ref{seds} are then used to create simulated SEDs and dervie broadband
photometric magnitudes. In this section we discuss the effects of
accretion and disc presence on the simulated observations.

@@

The left panels show accretion streams with an areal coverage of 10\%
and the left panels 1\%. As the areal coverage reduces the effective
temperature of the accretion hot spot increases, resulting in an
increase in accretion flux, and resulting shift to bluer wavelengths
of the peak flux. This can be seen clearly by comparing the left and
right panels of Figure \ref{acc_dom} As the rotation rate increases
the co-rotation radius also increases, as shown in Equation
\ref{inner_eq}, $R_{\rm inner}\propto \tau^{\frac{2}{3}}$. Therefore,
as Equations \ref{Lacc} and \ref{Tacc}, show, $L_{\rm acc}\propto 1-
\frac{R_*}{R_{\rm inner}}$ and $T_{\rm acc}\propto (L_{\rm
    acc}/A)^{\frac{1}{4}}$. Therefore, the temperature of the
accretion hot spot increases as the rotational period increases, due
to the increase in potential energy lost by the mass accreted. This
can be seen in Figure by comparing the left and right panels of Figure
\ref{acc_dom}. Perhaps the most important, albeit qualitative, result
shown is Figure \ref{acc_dom} is an insight into the accretion rate at
which the accretion blackbody flux dominates over the photospheric
flux. Figure \ref{acc_dom} shows that as the accretion rate raises
above \logmdot = $-9$ for systems with 1 or 10\% areal coverage, the
accretion flux dominates the emergent SED at both
periods. Effectively, the spectroscopic features of the photosphere
are veiled by the additional continuum accretion flux. Therefore for
reasonable coverages (1--10\%) and rotational periods (0.5--5 days)
the photospheric flux is effectively veiled by accretion flux for
accretion rates \logmdot$>-9$. It is also clear from Figure
\ref{acc_dom} that the magnitude becomes brighter and the colour bluer
(in terms of optical photometry) with increasing accretion rates as
expected \citep{gullbring_1998}. The impact on the photometry of BDD
systems is discussed in the next Section for individual stars and in
Section \ref{isochrones} for populations.

@@

Figure \ref{inner_inc} shows the SEDs for the viewing angles of 0, 27,
39, and 48$^{\circ}$. The top and bottom panels of Figure
\ref{inner_inc} show SEDs for models with a curved, as exemplified by
Figure \ref{flare_-7_193}, and a flat inner wall, as shown in Figure
\ref{flare_-12_183} respectively. The specific models chosen have been
selected to minimise flux differences between the two SEDs but have
distinct inner edges. The top panel has a mass of 0.04$M_{\odot}$,
\logmdot = -8, areal coverage of 1\%, a rotational period of 5 days
and an age of 10 Myrs. The bottom panel has a the same model but with
rotation period of 0.5 days and an age of 1 Myrs. Comparing the top
and bottom panel show that flux levels for the higher accretion rate,
with the larger and curved inner edge, vary less distinctly with
inclination angle than the systems with a vertical wall. The
difference in inclination dependence of the fluxes between the curved
and vertical inner walls is marginal. This is due to differences in
the stellar flux levels and disc structures.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/inner_inc.ps}
  \caption{Figure showing SEDs of systems with a curved and flat inner
    wall as the top and bottom panels respectively. The lines show the
    SEDs at viewing angles of 0, 27, 39, and 48$^{\circ}$.
\label{inner_inc}}
\end{figure*}

Figure \ref{inner_inc} shows only that for inclinations close to
face-on, the curved inner edge has SEDs with a stronger dependence on
the viewing angle compared to the model with a vertical inner
wall. This leads to a greater dependence of IR colour with
inclination.  Figure \ref{inner_mag} shows the difference in the
inclination dependence of the IR colour for large curved and smaller
vertical inner walls (shown as blue and black lines respectively), for
the systems shown in Figure \ref{inner_inc}. The dotted lines are
illustrative of the total change in $(J-K)_0$.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/inner_mag.ps}
  \caption{Figure showing the $(J-K)_0$ colour against the
    inclinations of 0, 27, 39, and 48$^{\circ}$, for the models shown
    in Figure \ref{inner_inc}. The models with the flat, vertical
    inner wall and curved inner edge are shown as the black and blue
    lines respectively. The horizontal dotted lines show the change in
    the colour index.\label{inner_mag}}
\end{figure*}

The $(J-K)_0$ colour of the higher accretion rate system changes less
as a function of inclination when compared to the lower accretion rate
system. A total change in $(J-K)_0$ of $\approx$ 0.6 and 0.05 are
found for the model with the flat and curved inner edge
respectively. This reduction in the change in IR colour as a function
of inclination for viewing angles close to face-on, is caused by the
larger (vertically) and convex inner wall in the extreme accreting
case. For the convex inner wall the effective emitting area one
observes does not change as much as a vertical wall as a function of
inclination, leading to this reduced dependence of IR colour on
inclination.

Therefore, for individual systems changes in the accretion rate and
disc structure lead to significant changes in magnitudes and
colours. The data presented in this section are for an isolated mass
and age. However, the trends presented are present throughout our
model grid for any given subset, and as such are representative. The
scatter or changes in magnitude in the individual systems will change
as a function of the remaining variables, but the dominant input
variables affecting simulated photometry are accretion rate and
inclination. For populations of stars these changes in magnitude and
colour act to scatter or spread a pre-MS BD locus in photometric
space.

@@

As discussed in Section \ref{disc_struct} the inner edge location is
correlated with inner edge temperature (albeit differently for the
typical and extreme accretors). As noted by \cite{meyer_1997} this
could lead to a correlation of IR excess with inner edge position. As
expected however, scattered caused by variations in inclination and
disc structure act to remove this correlation for populations.  The
correlation for an individual set of a systems, i.e. all variables
fixed except rotation rate, between IR colour and rotation rate is
preserved. However, as shown in Section \ref{disc_struct} the changes
caused by accretion rate and inclination angle are the most
significant, and act to remove any correlation between rotation rate
and IR colours for populations. In fact, for our grid the simulated
photometry shows no significant correlation between rotation rate and
IR colours. The work of \cite{meyer_1997} studied flat accreting disc,
therefore, as we have increased variation in disc structure by
applying vertical hydrostatic equilibrium it would be interesting to
see if any correlation appears for analytically defined disc
structures. We have run a set of parrallel models using analytical
disc structures and will publish the results in a subsequent
publication. Overall, as found in \cite{walker_2004}, the
dominant scattering effect for BD disc systems in an optical CMD
appears to be caused by accretion rate and inclination. Therefore for
derivation of parameters we explore the scatter in pre-MS BD locii as
a function of accretion rate and inclination.

@@

Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} shows CMDs in $M_V$
\& $(V-I)_0$ and $M_J$ \& $(J-K)_0$ resepctively. The left panels of
both figures shows stars classed as typical accretors with accretion
rates of \logmdot = $-$9, $-$10 and $-$11 \& $-$12, shown as blue,
black and red dots respectively in the top panels. The right panels of
Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} show systems with
extreme accretion rates, with \logmdot = $-$6, $-$7 and $-$8 shown as
blue, black and red dots respectively, in the top right panel. The
bottom panels then show the systems seperated into groups by
inclination. These groups are $\theta \leq 48^{\circ}$ as blue dots
(classed as face-on systems), $\theta> 56$ \& $64^{\circ}$ (classed as
the expected systems, as the expectation value of
$cos(\theta)=60^{\circ}$) as black dots and $\theta \geq 71^{\circ}$
(classed as edge-on systems) as red dots (for typical, bottom left,
and extreme, bottom right accretion rates). The left panels then show
naked stars isochrones (created from our grid of simulated photometry)
for 1 Myrs at accretion rates of \logmdot = $-$12 and $-$9, as solid and
dashed lines respectively (with coverage of 10\% and rotation rates of
5 days). Whereas the left hand panels show naked stars isochrones of
\logmdot = $-$6 and $-$9 as solid and dashed black lines respectively (with
coverage of 10\% and rotation rates of 5 days. The solid green line,
in the top panels, shows the 1 Myr isochrone of \cite{siess_2000}
adjusted to a distance of 250pc and an extinction of $A_V=2$ mag,
simulating a background population of CTTS stars.

@@

As can be seen in Figures \ref{age_mass_VVI} and \ref{age_mass_JJK}
current accretion and disc presence in a star and disc systems creates
a scatter in our simulated photometry indicative of a much larger
isochronal age or mass spread. Indeed, for many BDD systems even at
nominal accretion rates of \logmdot = $-$11 or $-$12, for our
simulations, the colours of these stars move significantly blueward of
the expected BD locus in a $M_V$, $(V-I)_0$ CMD and redward in a
$M_J$, $(J-K)_0$ CMD. As, in the case of typical accretors, the input
variables are in the range expected for a BD population, one could
reasonably expect observed true BD locii to show a similar scatter. It
is clear that even a wide photometric selection would not even include
all of the negligibly accreting systems at expected
inclinations. Whilst the derivation of masses and ages for these
objects will be difficult. Furthermore, for the higher accretion rates
of \logmdot = $-$9 or $-$10 the movement of the star within the CMD
will effectively move the star into the contamination region expected
for background CTTS or MS stars at a $(V-I)_0$ of $\leq$1.5 and $M_V$
12--10, and as such the star would not be included in a
photometrically selected BD sample. The solid green line, in the top
panels of Figures \ref{age_mass_VVI} and \ref{age_mass_JJK}, showing
the 1 Myr isochrone of \cite{siess_2000} at a distance of 250 pc and
extinction of $A_V=2$ mags shows that the BDD systems with higher
accretion rates could easily be confused for a background CTTS or MS
population. Indeed misclassification of a BDD system as a CTTS system
has already been revealed in \cite{white_2003}. For the extreme
accretors, as shown in the right panels of Figures \ref{age_mass_VVI}
and \ref{age_mass_JJK}, there is little chance of these objects being
classed, photometrically, as BD candidates or being assigned the
correct mass. Therefore, if one attempted to locate a population of
BDD systems with elevated or extreme accretion rates, target selection
would have to be placed at much brighter magnitudes, and bluer for
optical or redder for IR, colours.

@@

If one adopts the range of input parameters we have used (see Section
\ref{par_space} for justification), our simulated photometry shows,
qualitatively, that a coeval 1 Myr population of accreting BD stars
and BDD systems, with typical accretion rates and range of
inclinations, will exhibit a significant scatter in apparent
isochronal age. Furthermore, objects with typical (and extreme)
accretion rates are scattered sufficiently in CMD space to prohibit
their identification as pre-MS BDs. Indeed, these objects would not be
included in a photometrically selected sample of BDs, and as such are
unlikely to be assigned the correct masses or ages. The scatter from
the naked 1 Myr systems, generally, increases with increasing
accretion rate. It is important to note at this point that these
conclusions are qualitative, and obviously based on our
assumptions. However, the repercussion for isochronal age derivation
and sample selection in the BD regime could be profound. Indeed this
study has only included the effects of current or ongoing accretion
from a disc, it has neglected any effects of accretion, both past and
present, on the evolution of the central star. This past accretion
could also act to reduce the stars radius, accelerating contraction
\citep{tout_1999,siess_1999} and introducing additional scatter in a
coeval population proportional to the range in accretion rates
\citep[see][for full discussion]{mayne_2008}. In addition we have
shown that the typical accretion rates may scatter BDD systems into
the region of a CMD occupied by the CTTS or background MS locus.

@@

The observational cuts applied, shown as dashed lines, are from
\cite{luhman_2005b} a study of IC348, \cite{luhman_2008} a study of
$\sigma$ Orionis and \cite{gutermuth_2008} a study if the Taurus
region. In these cases the effects of extinction are either negligible
in the plotted colours, with values of $E([3.6]-[4.5])<0.04$ and
$E([4.5]-[5.8])<0.02$ for IC348 and $A_V \leq$4 mag \citep[which will
be negligible in the IRAC CoCoD,][]{allen_2004}, or the cuts
have been placed in intrinsic colour space as for $\sigma$
Orionis. The cuts from \cite{luhman_2005b}, \cite{luhman_2008} and
\cite{gutermuth_2008} appear as the top, top-middle and bottom-middle
panels. The lower panel shows a recent BDD candidate selection using
the $\alpha$ value from \citep{kennedy_2009}.

@@

Subsequently the SEDs of BDD systems with typical accretion rates and
associated discs change significantly from the assumed
underlying photospheric model flux, and therefore become difficult to
classify. In Section \ref{observables} we have shown that the BD
photosphere becomes veiled by the accretion flux for rates of
\logmdot$>-$9. The outer disc flaring observed in the BDD systems was
shown to cause occultation and a subsequent, sharp, fall in flux at an
inclination which decreases for more systems with higher accretion
rates. We have also shown that for extreme accretion rates the inner
wall increases in size and becomes convex in shape. We have also shown
that this leads to a reduced dependence on inclination (at
inclinations close to face-on) of the IR colours, compared to a
vertical wall.

>>>>

In surveys, particularly those attempting to discern the PMS disc
fraction this disentangling usually takes place using broad-band
photometric measures and cuts in colour-colour or colour-magnitude
space \citep{luhman_2005b,luhman_2008,gutermuth_2008}.  The important
quantity here is the separation in wavelength between the emission
peaks for the stellar and thermal disc components, as it is these
components which must be isolated. For BD disc (BDD hereafter) systems
it is likely that difficulties detecting the disc will be exacerbated
by the lower temperatures of the photosphere and therefore the smaller
separation in wavelength from the thermal disc emission component,
compared to CTTS systems. Therefore, to derive disc fractions for BD
populations we require detailed comparison models with which to guide
disc candidate selection.

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be indepedent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}. At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photopsheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may
well be a population of low-accretion-rate CTTS which are current
missing from the surveys.

@@

In summary the key required input variables to setup the model grid
are as follows: stellar age and mass (which are used to derive the
stellar flux) and accretion rate, areal coverage and rotation period
(which are used to determine the the accretion flux and temperature).
The disc parameters are set by a disc mass (expressed as a fraction of
the BD mass), and a power-law distribution for the disc surface
density.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a swift
convergence of the sublimated rim. The dust sublimation proceeds as
follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations. For a more
sophisticated, and chemically more complex, study of dust sublimation,
albeit for higher mass objects see \cite{kama_2009}.

@@

The disc surface density is conserved in the initial density
structure, $\Sigma ^{\beta - \alpha}$. The initial scaleheight ($h$)
and density ($\rho$) follow $h\propto r^{\beta}$
($h=h_0(r/R_*)^{\beta}$) and $\rho \propto r^{-\alpha}$ ($\rho
=\rho_0\frac{R_*}{r}exp(-\frac{1}{2}[z/h(r)]^2)$) respectively., where
$r$ and $z$ are the radial and vertical coordinates. The initial
values of $\alpha$ and $\beta$ were 2.1 and 1.1 respectively. The
values for $\alpha$ and $\beta$ were chosen to optimise resolution of
the vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{walker_2004}. The initial disc scaleheight at 100\,AU, $h(100)$,
was set to 25\,AU. As the simulation used vertical hydrostatic
equilibrium and dust sublimation, both the disc scaleheight and inner
edge location then evolved in the systems dependent on the input
parameters. In this section we discuss the structure of the discs in
terms of these two generated characteristics, i.e the disc scaleheight
and inner edge location.

@@

The resulting converged disc structures, as discussed in Section
\ref{seds} are then used to create simulated SEDs and derive broadband
photometric magnitudes. In this section we discuss the effects of
accretion and disc presence on the simulated observations.

@@

The left panels show accretion streams with an areal coverage of 10\%
and the left panels 1\%. As the areal coverage reduces the effective
temperature of the accretion hot spot increases, resulting in an
increase in accretion flux, and resulting shift to bluer wavelengths
of the peak flux. This can be seen clearly by comparing the left and
right panels of Figure \ref{acc_dom}. As the rotation rate increases
the co-rotation radius also increases, as shown in Equation
\ref{inner_eq}, $R_{\rm inner}\propto \tau^{\frac{2}{3}}$. Therefore,
as Equations \ref{Lacc} and \ref{Tacc} show $L_{\rm acc}\propto 1-
\frac{R_*}{R_{\rm inner}}$ and $T_{\rm acc}\propto (L_{\rm
  acc}/A)^{\frac{1}{4}}$, the temperature of the accretion hot spot
increases as the rotational period increases. This is due to the
increase in potential energy lost by the mass accreted. This can be
seen in Figure by comparing the left and right panels of Figure
\ref{acc_dom}. Perhaps the most important, albeit qualitative, result
shown is Figure \ref{acc_dom} is an insight into the accretion rate at
which the accretion blackbody flux dominates over the photospheric
flux. Figure \ref{acc_dom} shows that as the accretion rate raises
above \logmdot = $-9$ for systems with 1 or 10\% areal coverage, the
accretion flux dominates the emergent SED at both
periods. Effectively, the spectroscopic features of the photosphere
are veiled by the additional continuum accretion flux. Therefore for
reasonable coverages (1--10\%) and rotational periods (0.5--5 days)
the photospheric flux is effectively veiled by accretion flux for
accretion rates \logmdot$>-9$. It is also clear from Figure
\ref{acc_dom} that the magnitude becomes brighter and the colour bluer
(in terms of optical photometry) with increasing accretion rates as
expected \citep{gullbring_1998}. The impact on the photometry of BDD
systems is discussed in the next Section for individual stars and in
Section \ref{isochrones} for populations.

@@

For our model grid the inner edge shape clearly changes with
increasing flux levels and therefore accretion rate. This, for
individual systems will lead to the reduced dependence of the IR
colours on inclination as the edge moves from a vertical to a convex
boundary as found by \cite{isella_2005}. Practically, for populations
of stars however disentangling this effect from the influences of
various other parameters (such as changing disc scaleheights and
accretion rates) on the IR colours is difficult. Essentially, matching
flux levels between two systems whilst maintinaing a difference in
inner wall shape is impossible.

Therefore, for individual systems changes in the accretion rate and
disc structure lead to significant changes in magnitudes and
colours. The data presented in this section are for an isolated mass
and age. However, the trends described are evident throughout our
model grid for any given subset, and as such are representative. The
scatter or changes in magnitude in the individual systems will change
as a function of the remaining variables, but the dominant input
variables affecting simulated photometry are accretion rate and
inclination. For populations of stars these changes in magnitude and
colour act to scatter or spread a pre-MS BD locus in photometric
space.

@@

As discussed in Section \ref{disc_struct} the inner edge location is
correlated with inner edge temperature (albeit differently for the
typical and extreme accretors). As noted by \cite{meyer_1997} this
could lead to a correlation of IR excess with inner edge position. As
expected however, scatter caused by variations in inclination and disc
structure act to remove this correlation for populations.  The
correlation for an individual set of a systems, i.e. all variables
fixed except rotation rate, between IR colour and rotation rate is
preserved. However, as shown in Section \ref{disc_struct} the changes
caused by accretion rate and inclination angle are the most
significant, and act to remove any correlation between rotation rate
and IR colours for populations. In fact, for our grid the simulated
photometry shows no significant correlation between rotation rate and
IR colours. The work of \cite{meyer_1997} studied flat accreting disc,
therefore, as we have increased variation in disc structure by
applying vertical hydrostatic equilibrium it would be interesting to
see if any correlation appears for analytically defined disc
structures. We have run a set of parrallel models using analytical
disc structures and will publish the results in a subsequent
publication. Overall, as found in \cite{walker_2004}, the dominant
scattering effect for BD disc systems in an optical CMD appears to be
caused by accretion rate and inclination. Therefore for derivation of
parameters we explore the scatter in pre-MS BD locii as a function of
accretion rate and inclination.

@@

Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} shows CMDs in $M_V$
\& $(V-I)_0$ and $M_J$ \& $(J-K)_0$ respectively. The left panels of
both figures shows stars classed as typical accretors with accretion
rates of \logmdot = $-$9, $-$10 and $-$11 \& $-$12, shown as blue,
black and red dots respectively in the top panels. The right panels of
Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} show systems with
extreme accretion rates, with \logmdot = $-$6, $-$7 and $-$8 shown as
blue, black and red dots respectively, in the top right panel. The
bottom panels then show the systems seperated into groups by
inclination. These groups are $\theta \leq 48^{\circ}$ as blue dots
(classed as face-on systems), $\theta> 56$ \& $64^{\circ}$ (classed as
the expected systems, as the expectation value of
$cos(\theta)=60^{\circ}$) as black dots and $\theta \geq 71^{\circ}$
(classed as edge-on systems) as red dots (for typical, bottom left,
and extreme, bottom right accretion rates). The left panels then show
naked stars isochrones (created from our grid of simulated photometry)
for 1 Myrs at accretion rates of \logmdot = $-$12 and $-$9, as solid
and dashed lines respectively (with coverage of 10\% and rotation
rates of 5 days). Whereas the left hand panels show naked stars
isochrones of \logmdot = $-$6 and $-$9 as solid and dashed black lines
respectively (with coverage of 10\% and rotation rates of 5 days. The
solid green line, in the top panels, shows the 1 Myr isochrone of
\cite{siess_2000} adjusted to a distance of 250pc and an extinction of
$A_V=2$ mag, simulating a background population of CTTS stars.

@@

As can be seen in Figures \ref{age_mass_VVI} and \ref{age_mass_JJK}
current accretion and disc presence in a star and disc systems creates
a scatter in our simulated photometry indicative of a much larger
isochronal age or mass spread. Indeed, for many BDD systems even at
nominal accretion rates of \logmdot = $-$11 or $-$12, for our
simulations, the colours of these stars move significantly blueward of
the expected BD locus in a $M_V$, $(V-I)_0$ CMD and redward in a
$M_J$, $(J-K)_0$ CMD. As, in the case of typical accretors, the input
variables are in the range expected for a BD population, one could
reasonably expect observed true BD locii to show a similar scatter. It
is clear that even a wide photometric selection would not even include
all of the negligibly accreting systems at expected
inclinations. Whilst the derivation of masses and ages for these
objects will be difficult. Furthermore, for the higher accretion rates
of \logmdot = $-$9 or $-$10 the movement of the star within the CMD
will effectively shift the star into the contamination region expected
for background CTTS or MS stars at a $(V-I)_0$ of $\leq$1.5 and $M_V$
12--10, and as such the star would not be included in a
photometrically selected BD sample. The solid green line, in the top
panels of Figures \ref{age_mass_VVI} and \ref{age_mass_JJK}, showing
the 1 Myr isochrone of \cite{siess_2000} at a distance of 250 pc and
extinction of $A_V=2$ mags shows that the BDD systems with higher
accretion rates could easily be confused for a background CTTS or MS
population. Indeed misclassification of a BDD system as a CTTS system
has already been revealed in \cite{white_2003}. For the extreme
accretors, as shown in the right panels of Figures \ref{age_mass_VVI}
and \ref{age_mass_JJK}, there is little chance of these objects being
classed, photometrically, as BD candidates or being assigned the
correct mass. Therefore, if one attempted to locate a population of
BDD systems with elevated or extreme accretion rates, target selection
would have to be placed at much brighter magnitudes, and bluer for
optical or redder for IR, colours.

@@

If one adopts the range of input parameters we have used (see Section
\ref{par_space} for justification), our simulated photometry shows,
qualitatively, that a coeval 1 Myr population of accreting BD stars
and BDD systems, with typical accretion rates and range of
inclinations, will exhibit a significant scatter in apparent
isochronal age. Furthermore, objects with typical (and extreme)
accretion rates are scattered sufficiently in CMD space to prohibit
their identification as pre-MS BDs. Indeed, these objects would not be
included in a photometrically selected sample of BDs, and as such are
unlikely to be assigned the correct masses or ages. The scatter from
the naked 1 Myr systems, generally, increases with increasing
accretion rate. It is important to note at this point that these
conclusions are qualitative, and obviously based on our
assumptions. However, the repercussions for isochronal age derivation
and sample selection in the BD regime could be profound. Indeed this
study has only included the effects of current or ongoing accretion
from a disc, it has neglected any effects of accretion, both past and
present, on the evolution of the central star. This past accretion
could also act to reduce the stars radius, accelerating contraction
\citep{tout_1999,siess_1999} and introducing additional scatter in a
coeval population proportional to the range in accretion rates
\citep[see][for full discussion]{mayne_2008}. In addition we have
shown that the typical accretion rates may scatter BDD systems into
the region of a CMD occupied by the CTTS or background MS locus.

@@

The observational cuts applied, shown as dashed lines, are from
\cite{luhman_2005b} a study of IC348, \cite{luhman_2008} a study of
$\sigma$ Orionis and \cite{gutermuth_2008} a study of the Taurus
region. In these cases the effects of extinction are either negligible
in the plotted colours, with values of $E([3.6]-[4.5])<0.04$ and
$E([4.5]-[5.8])<0.02$ for IC348 and $A_V \leq$4 mag \citep[which will
be negligible in the IRAC CoCoD,][]{allen_2004}, or the cuts
have been placed in intrinsic colour space as for $\sigma$
Orionis. The cuts from \cite{luhman_2005b}, \cite{luhman_2008} and
\cite{gutermuth_2008} appear as the top, top-middle and bottom-middle
panels. The lower panel shows a recent BDD candidate selection using
the $\alpha$ value from \citep{kennedy_2009}.

@@

Subsequently the SEDs of BDD systems with typical accretion rates and
associated discs change significantly from the assumed underlying
photospheric model flux, and therefore become difficult to
classify. In Section \ref{observables} we have shown that the BD
photosphere becomes veiled by the accretion flux for rates of
\logmdot$>-$9. The outer disc flaring observed in the BDD systems was
shown to cause occultation and a subsequent, sharp, fall in flux at an
inclination which decreases for more systems with higher accretion
rates. We have also shown that for extreme accretion rates the inner
wall increases in size and becomes convex in shape.
